\thetitle

### 2021-03-02 writing/challenges.tex
a bit more work on challenges

However, three major challenges remain:  the isolation of the effect of
blame on the rational programmer across three different mechanisms; the
collection of a large number of representative debugging scenarios;  and a
proper sampling strategy for the huge space of possibilities. A
coincidental challenge is the disparity of the implementations of
gradually typed languages. To eliminate this variable from the comparison,
this paper uses Racket, which is thus far the only language in which all
three major soundness mechanisms are available in a reasonably robust and
directly comparable manner; for the implementation of the transient
semantics, see the parallel submission of an experience
report~\cite{ttt21}.

The {\em first major challenge\/} stems from the interplay between blame and the
run-time checks of a gradual type systems. When a check fails the Natural
and Transient semantics assign blame and discard  the information in the
exception from the run-time check. Nevertheless, the exception
information may be equally or even more useful to the rational programmer
than blame. If this is the case blame may not play a critical role for the
rational programmer. Indeed, precisely because they do not
account for confounding factors, ~\citet{lksfd-popl-2020} cannot draw any
conclusions about the pragmatics of blame despite advertisements for the
opposite. Their experiment can only conclude that so-called blame-shifting
works but not that this is because of blame.

One way to isolate blame from this confounding factor is to {\em equip the
rational programmer with modes \/} that represent the different types of
information the rational programmer takes into account when debugging a
scenario. Specifically, the rational programmer comes with a blame mode and
an exception mode. Only if the former succeeds debugging a program while
the latter fails, it is safe to conclude that blame is indeed necessary
for the rational programmer. Put differently, the exception mode serves as
the {\em baseline\/} for blame's value; if the programmer in this mode
``beats'' the  blame one, a blame assignment mechanism might be useless.

Besides the run-time checks, pure luck might also amplify the usefulness
of blame. Thus a proper evaluation of blame needs to factor out any random
effects. A random rational programmer mode provides the necessary baseline
using the same ingredients as for exceptions.

Modes also offer a solution to a derivative challenge; reconciling the
differences among the blame assignment mechanisms of the three semantic variants.
While Natural assigns blame to {\em one\/} component, Transient assigns
blame to a sequence of components. The Erasure semantics does not blame a
component {\it per se\/}, but it comes with an exception location and a
stack trace, which implicitly suggests parties to blame.  The different
strategies greatly affect the behavior of the rational programmer (and
real programmers, too).

With modes, different blame strategies map to different ways the rational
programmer  interprets blame.  For instance, one Transient mode may pick
the oldest element of a blame set as the next component to add types
because it corresponds to the earliest point in the program's evaluation
that can discover an impedance mismatch.  Another mode may opt to treat
the sequence as a stack and add types to its newest element.  When both
modes are equally successful in locating an impedance mismatch,
measuring the rational programmer's effort may answer which mode is the
most effective.

>>>>

However, three major challenges remain:  comparing the effect of
blame on the rational programmer across three different mechanisms; the
collection of a large number of representative debugging scenarios;  and a
proper sampling strategy for the huge space of possibilities. A
coincidental challenge is the disparity of the implementations of
gradually typed languages. To eliminate this variable from the comparison,
this paper uses Racket, which is thus far the only language in which all
three major soundness mechanisms are available in a reasonably robust and
directly comparable manner; for the implementation of the transient
semantics, see the parallel submission of an experience
report~\cite{ttt21}.

The {\em first challenge\/} stems from the differences between the blame
assignment mechanisms of the three semantic variants.  While Natural
assigns blame to {\em one\/} component, Transient assigns blame to a
sequence of components. The Erasure semantics does not blame a component
{\it per se\/}, but it comes with an exception location and a stack trace,
which implicitly suggests parties to blame.  The different strategies
greatly affect the behavior of the rational programmer (and real
programmers, too).

One way to reconcile these differences is to {\em equip the rational
programmer with modes \/} that represent the different types of
information the rational programmer takes into account when debugging a
scenario. Intuitively, with the common vocabulary of modes, different
blame strategies become different ways the rational programmer  interprets
blame.  For instance, one Transient mode may pick the oldest element of a
blame set as the next component to add types because it corresponds to the
earliest point in the program's evaluation that can discover an impedance
mismatch.  Another mode may opt to treat the sequence as a stack and add
types to its newest element.  When both modes are equally successful in
locating an impedance mismatch, measuring the rational programmer's effort
may answer which mode is the most effective.

However, attributing the success of the rational programmer to this or
that blame mechanism demands a careful look at the interplay between blame
and the run-time checks of each gradual type system. When a check fails
the Natural and Transient semantics assign blame and discard  the
information in the exception from the run-time check. Nevertheless, the
exception information may be equally or even more useful to the rational
programmer than blame. If this is the case blame may not play a critical
role for the rational programmer. Indeed, precisely because they do not
account for confounding factors, ~\citet{lksfd-popl-2020} cannot draw any
conclusions about the pragmatics of blame despite advertisements for the
opposite. Their experiment can only conclude that so-called blame-shifting
works but not that this is because of blame.

Modes offer a uniform way to compare the different semantics and  isolate
blame from the effect of the semantics's runtime checks. Specifically, the Natural and
Transient rational programmer come with a blame mode and an exception
mode. Only if the former succeeds debugging a program while the latter
fails, it is safe to conclude that blame is indeed necessary for the
rational programmer. Put differently, the exception mode serves as the
{\em baseline\/} for blame's value within a given semantics; if the
programmer in this mode ``beats'' the  blame one, a blame assignment
mechanism might be useless.

Besides the run-time checks, pure luck might also amplify the usefulness
of blame. Thus a proper evaluation of blame needs to factor out any random
effects. A random rational programmer mode provides yet another necessary
baseline using the same ingredients as for exceptions.
